---
title: "Assessing food web stability with the fwstability R-package."
author: "D.S.W. de Jonge"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fwstability)
```

This vignette explains how you can use the fwstability package to assess food web stability. There are basically three sections: 1) how to prepare your food web model for stability analysis, 2) how to conduct stability analysis, and 3) how to assess your food web for characteristics related to stability.

# 1. Prepare a food web model for stability analysis

# 2. Stability analysis
For now we prepare a fake Jacobian matrix (JM).
```{r JM}
JM <- matrix(data = c(
  0,     0.2, 0,   0.4,  -0.1,
  -0.05, 0,   0.1, -0.5, 0.2,
  -0.2,  0.1, 0,   0.1,  -0.6,
  0.4,  -0.3, -0.1, 0,   0.3,
  0.1,  -0.2, 0.1, -0.5, 0
  ), nrow = 5, ncol = 5, byrow = TRUE
)
```

The stability of this Jacobian matrix is calculated as the maximum real value of the eigenvalues. A negative value indicates the food web is stable. 
```{r}
getStability(JM)
```


# 3. Analysize food web for (de)stabilizing characteristics
