---
title: "Assessing food web stability with the fwstability R-package."
author: "D.S.W. de Jonge"
date: "`r date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Quick start
This vignette explains how you can use the fwstability package to assess food web stability. There are basically four sections:  
1) how to prepare your food web model for stability analysis,  
2) how to conduct stability analysis,  
3) how to assess your food web for characteristics related to stability, and  
4) examples with food webs from literature.  
These sections contain details about all the features of the package and the scientific background. Reading this is highly recommended so that you can interpret your results correctly. This quick start section will give you a general idea about the usage of the package.

First, we load the package.
```{r setup}
library(fwstability)
```

# 1. Preparing a food web model for stability analysis
## A linear inverse model
A linear inverse food web model (LIM) can be prepared with the R-package LIM (https://cran.r-project.org/web/packages/LIM/index.html). This example food web model is the same food web model as in the LIM vignette, so please refer to this document to understand the concept and set-up of a LIM.
```{r LIM}
library(LIM)
lim <- Setup(system.file("extdata", "foodweb.lim", package = "fwstability"))
FM <- Flowmatrix(lim)

print(lim$Components)
print(lim$Externals)
print(FM)
```

This food web model consists of three compartments of which the biomass is measured; Plants, Animals, and Detritus, and two external compartments with infinite biomass; CO2 and export. The unit of biomass is g C m-2 and the unit of flow flux is g C m-2 d-1. Plants take up CO2 and are consumed by Animals. Animals also consume Detritus. Both Plants and Animals deposit carbon into the Detritus compartment through natural mortality and defecation. Animals respire into the CO2 compartment and export carbon from the system, for example when they are predated upon by external animals. Detritus decomposes and thereby releases carbon into the CO2 pool. 

Some of these flows are constrained by measured values. Therefore, linear programming can be used to find the parsimonious solution of all the flows (see LIM vignette). The quantified flows are stored in FM. 

Based on the final flow values and the constraints defined in the model it is possible to calculate the assimilation and growth efficiencies of animals and plants. Assimilation efficiency is 1 - (defecation / ingestion) and assumed 1 for Plants. Growth efficiency is defined in model for Animals and assumed 1 for Plants. The vectors containing biomass, assimilation and growth efficiencies must be named for usage in the getJacobian function. If assimilation and growth efficiencies are not specified in the model, or cannot be calculated from the model, you can estimate a value or try to find a value from literature.

```{r}
# Order Plant, Animal, Detritus. Just like the FM.
AE <- c(1, 1 - (FM[2,3] / sum(FM[,2])), NA)
GE <- c(1, 0.3, NA)
BM <- lim$Components$val
names(AE) <- rownames(FM)[1:3]
names(GE) <- rownames(FM)[1:3]
names(BM) <- rownames(FM)[1:3]
```

The food web data and calculated flows can be used to calculate a Jacobian matrix with the function getJacobian(). It is important to tell the function which compartments to regard as dead compartments (in this case detritus) and which compartments are external.
```{r getJacobian}
JM <- getJacobian(
  FM = FM,
  BM = BM,
  AE = AE,
  GE = GE,
  dead = c("DET"),
  externals = c("CO2", "EXP")
)

print(JM)
```

The interaction strengths have the unit per time. In our food web model the flows were g C m-2 d-1 and biomass g C m-2, so the interaction strengths have the unit d-1.

From the interaction strengths in the Jacobian matrix we observe that Animals have a negative effect on the growth of Plants, and Plants have a positive effect on Animals, which makes sense because Plants are eaten by Animals. Detritus for this same reason has a positive effect on Animals. Both Plants and Animals have a positive effect on detritus, despite the fact that Detritus is consumed by Animals. This is caused by the fact that Animals overall deposit more Detritus than they take up. This highlights the importance of including Detritus as a dead compartment in the function. If we would leave out this argument the Detritus compartment would be viewed as a living compartment which can actively consume other compartments. In that case the deposition of feces would be regarded as predation on the animals by detritus, which makes no sense.  

In the getJacobian function, the effect of a predator on its prey ($\alpha_{ij}$) is calculated as:

$$\alpha_{ij}=-\frac{F_{ij}}{x{j}}$$
and the effect of a prey on its predator($\alpha_{ji}$) is calculated as:

$$\alpha_{ji}=\frac{a_jp_jF_{ij}}{x{i}}$$
where $F_{ij}$ is the flux between resource population $i$ and consumer population $j$, $a_j$ is the assimilation efficiency of the consumer, $p_j$ is the production efficiency of the consumer, and $x_{i,j}$ is the biomass of the respective population (de Ruiter et al., 1995; Neutel et al., 2002).

The effect of a population $j$ on detritus ($\alpha_{dj}$) is calculated as:

$$\alpha_{dj}=\frac{1}{x_j}(F_{jd} - F_{dj} + \sum_{k=1}^n (1-a_k)F_{jk})$$
where $n$ is the total number of compartments in the food web (Neutel & Thorne, 2014).

The diagonal of a Jacobian matrix represents intraspecific interactions. The default diagonal of the function getJacobian() is to set all values to zero. Neutel & Thorne (2014) show that an all-zero diagonal (with exception of detritus flows) has the biological meaning of a tipping point where any loss of self-dampening effect or intraspecific competition will cause the system to collapse. Diagonal values can also be calculated from the energy-flux model. Diagonal values for fauna are found as their non-predatory mortality (Neutel & Thorne, 2014):
$$\alpha_{ii}=-\frac{\sum_{j=1}^n a_i p_i F_{ji}-\sum_{j=1}^n F_{ij}}{x_i}$$
and for detritus as:
$$\alpha_{dd}=-\frac{1}{x_d} \sum_{j=1}^n (a_j F_{dj})$$

The Jacobian matrix we calculated earlier had the default all-zero diagonal. You can also decide to calculate diagonal values from the flux values as described by the equations above. In order to do this, natural mortality rates (biomass per unit time) are needed. There are several ways to obtain these values:  

1) Sometimes your food web model will specifically include a natural mortality flow. You can simply extract this information from your model into a named vector and use it as input.  
2) You can multiple the biomass by the inverse of the lifespan. The inverse of the lifespan represents the fraction of biomass that will disappear due to natural mortality on a certain time scale. For example, the biomass of animals in the example food web model is 50 g C m-2 and the natural lifespan of these animals is assumed to be 30 days. Then the natural mortality rate is 50 * 1/30 = 1.67 g C m-2 d-1. This lifespan should NOT include mortality due to predation, and be a good representative for the whole compartment.  

To calculate diagonal values based on the model we include the argument "diagonal" and set the value to the string "model". In our model natural mortality flows are already specified, so we can extract this information and use it as input for the mortality rates argument "MR". The mortality rate for Detritus is set to NA.

```{r getJacobian_diagonal}
MR <- c(FM["PL","DET"], FM["AN", "EXP"], NA)
names(MR) <- rownames(FM)[1:3]

JM2 <- getJacobian(
  FM = FM,
  BM = BM,
  AE = AE,
  GE = GE,
  dead = c("DET"),
  externals = c("CO2", "EXP"),
  diagonal = "model",
  MR = MR
)

print(JM2)
```

Some notes:
- The inclusion of biomasses in the LIM is not required to be able to solve the model. However, it is required that all internal compartments have quantified biomasses to calculate the Jacobian matrix. If there are compartments that do not have biomasses in the LIM, they must either be quantified anyway or excluded from the getJacobian() function by including them as 'externals'.
- The diagonal values calculated from the model for non-dead compartments represent the UPPER bound (in absolute value). This means we assume maximum self-dampening effects of populations. Jacobian matrixes with model-based diagonals are therefore often stable. If you want to compare stability between two food web models that are initially both stable, you can redefine stability as the minimum fraction of natural mortality needed to stabilize the matrix (see below 2. Stability analysis).

## A dynamic model
A dynamic food web model is a set of differential equations that describe the change in biomass of each food web compartment based on the biomass of other compartments and potentially other factors. It is common to use a set of (generalized) Lotka-Volterra equations for your food web model. The form of a generalized Lotka-Volteraa equation is:
$$\frac{dx_i}{dt} = x_i (r_i + \sum_{j=1}^n c_{ij}x_j) $$
where $r_i$ is the natural per capita growth rate and $c_ij$ is the coefficient of the interaction i.e. the effect of population $j$ on population $i$. 

Lets create a dynamic model based on the same food web structure and data as the linear inverse model described above. Plants have a natural growth rate as they take up CO2 for primary production. Furthermore, plants are negatively impacted by animals as they are preyed upon. Plants have a negative effect on their own population growth, and are not influenced by detritus. Growth of animals is dependen on the feeding on plants. Animals, like plants, have a negative effect on their own population growth. Detritus have natural decline by decomposition, but are positively affected by animal defecation and plant waste.

The deSolve package is great for solving a system of orginary differential equations. You can write your model in R, plug in parameters, and simulate the change over time in your ecosystem.
```{r Lotka-Volterra}
library(deSolve)
library(ggplot2)

fwnames <- c("PLANT", "ANIMAL", "DETRITUS")

# Define the set of differential equations
LotkaVolterra <- function (Time, State, Pars) {
  with(as.list(c(State, Pars)), {
    dP = PLANT * (r1 - c11 * PLANT - c12 * ANIMAL)
    dA = ANIMAL * (-r2 + c21 * PLANT - c22 * ANIMAL - c23 * DET)
    dD = DET * (-r3 + c31 * PLANT + c32 * ANIMAL - c33 * DET)
    return(list(c(dP, dA, dD)))
  })
}

# Parameters to initialize model
#Pars <- c(alpha = 2, beta = 0.5, gamma = 0.2, delta = 0.6)
Pars <- c(
  r1 = 10, r2 = 1, r3 = 1,
  c11 = 0.05, c12 = 0.5,
  c21 = 7, c22 = 0.5, c23 = 0.5,
  c31 = 0.75, c32 = 3, c33 = 2
)
# Initial biomass of compartments
State <- c(PLANT = 10, ANIMAL = 5, DET = 10)
# Time to run the simulation
Time <- seq(0, 50, by = 1)

# Run and plot simulation
simulation <- ode(func = LotkaVolterra, y = State, parms = Pars, times = Time)
plot <- ggplot(as.data.frame(simulation)) +
  geom_line(aes(x = time, y = PLANT), colour = "green") +
  geom_line(aes(x = time, y = ANIMAL), colour = "blue") +
  geom_line(aes(x = time, y = DET), colour = "brown")

```

Mathematically speaking, the Jacobian matrix contains elements $\alpha_{ij}$ which are the first-order partial derivatives of each differential function valued with initial state biomasses. The rootSolve package numerically estimates these values by perturbing the initial state values.

```{r}
newState <- simulation[50,2:4]
JM <- jacobian.full(y=newState, func=LotkaVolterra, parms = Pars)
rownames(JM) <- fwnames ; colnames(JM) <- fwnames
print(JM)
```

This Jacobian matrix already contains diagonal values. The stability can be found with this complete Jacobian matrix.
```{r}
getStability(JM)
```

# 2. Stability analysis
A Jacobian matrix is only stable if all real parts of the eigenvalues are negative. Therefore, the stability of a Jacobian matrix can be measured as the maximum real value of the eigenvalues. A negative value indicates a stable matrix, a positive value indicates an unstable matrix.
```{r}
# Using the default 'eigenvalue' method.
# Jacobian matrix with all-zero diagonal
getStability(JM)
# Jacobian matrix with calculated values on the diagonal
getStability(JM2)
```

The first Jacobian matrix with an all-zero diagonal is an unstable matrix, whereas the second Jacobian matrix with a diagonal calculated from the model is indeed stable. This highlights the importance of thinking about how you want to quantify your diagonal. 
If you use different methods to quantify the diagonal the stability of different models are are not comparable.

There is a second stability measurement that is commonly used, that allows comparison of food web models that are initially stable. In this case the minimum fraction of self-dampening needed to stabilize the Jacobian matrix is used as measure. For example, two Jacobian matrices with a diagonal calculated from the model are stable i.e. the maximum real part of the eigenvalues is negative. The diagonal of both Jacobian matrices contain the upper bound values representing the maximum possible self-dampening effect possible. If you start reducing the self-dampening effect, at some point the Jacobian matrix will become unstable. If the first matrix becomes unstable with 75% of its original self-dampening and the second becomes unstable at 50% of its original self-dampening, we can conclude that the second matrix is more stable than the first one. The second matrix requires a lot less intra-specific feedback to dampen the destabilizing effects of interspecific interactions.

This concept is implemented in the getStability() function using the 'scalar' method. For this method the additional argument mortalities is required, and 'dead' is required only if there are indeed dead compartments.
```{r}
getStability(JM2, method = "scalar", 
             mortalities = MR/BM, dead = "DET")
```

In this case, the diagonal value for detritus is calculated from the model (see 1. Preparing a food web model for stability analysis). If we then scale the natural mortality rates of non-dead compartments we see that the minimum fraction natural mortality needed to stabilize the matrix is 0.45. This means that at least 45% of the maximum self-dampening effect of populations is needed in order to buffer pertubations to the whole system.

# 3. Analysize food web for (de)stabilizing characteristics

# 4. Examples from literature

## de Ruiter et al. (1995) Energetics, Patterns of Interaction Strenghts, and Stability in Real Ecosystems. Science, 269 (5228), 1257 - 1260

In the Science paper by de Ruiter et al. (1995) the stability of seven food webs is assessed. The soil food web of the Lovinkhoeve Experimental Farm is singled out and a energy-flux model is provided including biomasses (kg ha-1), specific death rates (yr-1), assimilation efficiencies, and growth efficiencies for all compartments. 

```{r model_data}
# Get food web data on the compartments
foodweb_data <- list(
  # Compartment names
  compartment = c(
    "Detritus", 
    "Roots", 
    "Fungi",
    "Bacteria", 
    "Phytophagous_nematodes", 
    "Collembola", 
    "Cryptostigmatic_mites",
    "Noncryptostigmatic_mites",
    "Fungivorous_nematodes",
    "Enchytraeids",
    "Bacteriophagous_nematodes",
    "Flagellates",
    "Bacteriophagous_mites",
    "Amoebae",
    "Predatory_nematodes",
    "Nematophagous_mites",
    "Predatory_collembola",
    "Predatory_mites"
  ),
  # Biomasses(kg ha-1)
  BM = c(
    2500,
    300,
    2.13,
    227.5,
    0.19,
    0.47,
    0.01,
    0.02,
    0.08,
    0.43,
    0.30,
    0.53,
    0.001,
    11.53,
    0.06,
    0.004,
    0.03,
    0.0635
  ),
  MR = c(
    NA,
    1.00,
    1.20,
    1.20,
    1.08,
    1.84,
    1.20,
    1.84,
    1.92,
    5.00,
    2.68,
    6.00,
    1.84,
    6.00,
    3.00,
    1.84,
    1.84,
    1.84
  ),
  AE = c(
    NA,
    NA,
    1.00,
    1.00,
    0.25,
    0.50,
    0.50,
    0.50,
    0.38,
    0.25,
    0.60,
    0.95,
    0.50,
    0.95,
    0.50,
    0.90,
    0.50,
    0.60
  ),
  GE = c(
    NA,
    NA,
    0.30,
    0.30,
    0.37,
    0.35,
    0.35,
    0.35,
    0.37,
    0.40,
    0.37,
    0.40,
    0.35,
    0.40,
    0.37,
    0.35,
    0.35,
    0.35
  )
)
names(foodweb_data$BM) <- foodweb_data$compartment
names(foodweb_data$MR) <- foodweb_data$compartment
names(foodweb_data$AE) <- foodweb_data$compartment
names(foodweb_data$GE) <- foodweb_data$compartment

FM <- matrix(
  0,
  nrow = length(foodweb_data$compartment), 
  ncol = length(foodweb_data$compartment),
  byrow = T
)
if(T){
  FM[1,3] <- 40.8
  FM[1,4] <- 1606
  FM[2,5] <- 7.44
  FM[3,6] <- 6.65
  FM[3,7] <- 0.08
  FM[3,8] <- 0.34
  FM[3,9] <- 2.51
  FM[3,10] <- 0.09
  FM[1,10] <- 11.1
  FM[4,10] <- 10.1
  FM[4,11] <- 6.89
  FM[4,12] <- 9.46
  FM[4,13] <- 0.01
  FM[4,14] <- 182
  FM[11,15] <- 0.42
  FM[4,15] <- 0.33
  FM[12,15] <- 0.008
  FM[12,14] <- 0.42
  FM[14,15] <- 0.17
  FM[5,18] <- 0.06
  FM[5,17] <- 0.13
  FM[5,16] <- 0.01
  FM[5,15] <- 0.28
  FM[6,18] <- 0.31
  FM[7,18] <- 0.005
  FM[8,18] <- 0.015
  FM[9,18] <- 0.03
  FM[9,17] <- 0.05
  FM[9,16] <- 0.004
  FM[9,15] <- 0.11
  FM[13,18] <- 0.003
  FM[11,18] <- 0.10
  FM[11,17] <- 0.20
  FM[11,16] <- 0.014
  FM[15,18] <- 0.02
  FM[15,17] <- 0.04
  FM[15,16] <- 0.003
  FM[16,18] <- 0.002
  FM[17,18] <- 0.02
}

rownames(FM) <- foodweb_data$compartment
colnames(FM) <- foodweb_data$compartment
```

Now we use this data to calculate interaction strength values to get a Jacobian matrix. The authors state that intragroup interferences i.e. the values for the diagonal of the Jacobian matrix, could not be established empirically. Therefore, the diagonal is set to zero for now. In the original paper the interaction strengths related to the detritus compartment were not calculated with a function specific for detritus. Therefore, the results from the paper can be duplicated by NOT defining  detritus as a dead compartment.

```{r}
JM <- getJacobian(
  FM = FM,
  BM = foodweb_data$BM,
  AE = foodweb_data$AE,
  GE = foodweb_data$GE
)
```

Taking into account new insights from Neutel & Thorne (2014) the effects of trophic groups on detritical compartments can be calculated with an altered equation. To calculate the interaction strengths based on this other equation we have to include detritus as a dead compartment.

```{r}
JM <- getJacobian(
  FM = FM,
  BM = foodweb_data$BM,
  AE = foodweb_data$AE,
  GE = foodweb_data$GE,
  dead = "Detritus"
)
```

The stability of the food web can now be found by finding the maximum real part of the eigenvalues. Remember our Jacobian matrix has an all-zero diagonal.
```{r}
getStability(JM)
```

It is also possible to find stability as the scalar of the natural mortality rates needed to produce a stable matrix.
```{r}
getStability(JM, method = "scalar", 
             mortalities = foodweb_data$MR, dead = "Detritus")
```

Both methods are valid, altough it is clear that the actual values differ. Therefore, it is important to know what method has been chosen to find stability and to interpret it correctly.

It is also possible to calculate values for the diagonal instead of setting them to zero.
```{r}
JM <- getJacobian(
  FM = FM,
  BM = foodweb_data$BM,
  AE = foodweb_data$AE,
  GE = foodweb_data$GE,
  diagonal = "model",
  dead = "Detritus",
  MR = foodweb_data$MR
)

getStability(JM)
```


